Differential Expression - MSSM-Penn-Pitt - DLPFC - Sex

### 2020-07-30 individual-cohorts/MSSM-Penn-Pitt_DLPFC_DE.Rmd
test Sex

---
title: "Differential Expression - MSSM-Penn-Pitt - DLPFC - Dx_Sex"
author: "Kelsey Montgomery, Gabriel Hoffman, Thanneer Perumal"
date: "`r date()`"
output: html_document
---

@@

knit2synapse::createAndKnitToFolderEntity(file = "individual-cohorts/MSSM-Penn-Pitt_DLPFC_DE.Rmd",
                                 parentId ="syn22043922",
                                 folderName = "Differential Expression - Dx - Sex")
```

@@

```{r synapse.parameters, include=FALSE, cache=FALSE}
parentId = 'syn22043922';
activityName = 'Differential Expression';
activityDescription = 'Differential Expression - Dx - Sex';

@@

# Test Dx-Sex
COVARIATES <- COVARIATES %>%
  tibble::rownames_to_column(var = "samples") %>% 
  tidyr::unite("Dx_Sex",
               dplyr::all_of(c("Dx", "Reported_Gender")),
               remove = TRUE) %>% 
  dplyr::mutate(`Dx_Sex` = as.factor(`Dx_Sex`)) %>% 
  tibble::column_to_rownames(var = "samples")

# Set covariates to adjust for from model selection
adjust.covars <- c("Dx_Sex", "ageOfDeath", "RIN2", "RIN", "IntronicRate", "IntragenicRate")
random_effect <- c("Institution")
```

### Normalisation - Dx_Sex
1. Dx_Sex is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis)

@@

### Residual calculation
Calculate weighted residuals and add back "Dx_Sex" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx_Sex", colnames(ADJUSTED.FIT$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(ADJUSTED.FIT$design[,varsToAddIn])
```

@@

```{r diffExp, fig.height=10, fig.width=15}
L1 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx_SexControl_Female", "Dx_SexControl_Male"))
L2 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx_SexSCZ_Female", "Dx_SexSCZ_Male"))
L3 = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("Dx_SexOther_Female", "Dx_SexOther_Male"))
# Manually set comparison of SCZ-Other to Control
L = cbind(L1, L2, L3, 
          data.frame(L4 = c(1,0,-0.5, 0, -0.5, 0, 0, 0, 0, 0)),
          data.frame(L5 = c(0, 1, 0, -0.5, 0, -0.5, 0, 0, 0, 0)))

contrast <- c(glue::glue_collapse(names(L1[L1 != 0]), "-"), 
              glue::glue_collapse(names(L2[L2 != 0]), "-"),
              glue::glue_collapse(names(L3[L3 != 0]), "-"),
              glue::glue("Dx_SexControl_Female-Dx_SexOther_FemaleDx_SexSCZ_Female"),
              glue::glue("Dx_SexControl_Male-Dx_SexOther_MaleDx_SexSCZ_Male")
)

@@

# Get differential expression
DE = lapply(glue::glue("L{1:5}"), function(i, FIT){
  topTable(FIT, coef=i, number = Inf) %>%
    rownameToFirstColumn('gene_id') %>% 
    left_join(backgroundGenes)
}, FIT.CONTR) 
names(DE) <- contrast

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('Dx_Sex','',Comparison),
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, into = c('ref.state','to.state'), sep = '-') %>%
  tidyr::unite(Comparison, `ref.state`, `to.state`, sep = '_vs_') %>%
  left_join(Ensemble2HGNC) %>%
  left_join(gene_lengths)
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

@@

# Code
CODE <- Folder(name = "Differential Expression - Dx - Sex", parentId = parentId)
CODE <- synStore(CODE)

# Store residual gene expression for network analysis
RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_DLPFC_Dx_Sex_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
nEXP_OBJ = File('CMC_DLPFC_Dx_Sex_netResidualExpression.tsv', 
                name = 'Dx_Sex Normalised, covariates removed residual expression (for network analysis)', 
                parentId = CODE$properties$id)
nEXP_OBJ = synStore(nEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(nEXP_OBJ, annotations = all.annotations)

# Store differential expression results
write.table(DE, file = 'CMC_DLPFC_Dx_Sex_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('CMC_DLPFC_Dx_Sex_DiffExpression.tsv', 
                name = 'Differential Expression Results (Dx_Sex)', 
                parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(DEXP_OBJ, annotations = all.annotations)

# Store adjusted fit linear mixed model object as RDS - Dx_Sex
saveRDS(ADJUSTED.FIT, file = 'CMC_DLPFC_Dx_Sex_dream-output.rds')
ADJ_OBJ = File('CMC_DLPFC_Dx_Sex_dream-output.rds',
               name = 'Dream Output object - Dx_Sex', 
               parentId = CODE$properties$id)
ADJ_OBJ = synStore(ADJ_OBJ, used = ALL_USED_IDs, activityName = activityName,
                   executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(ADJ_OBJ, annotations = all.annotations)
```

>>>>

---
title: "Differential Expression - MSSM-Penn-Pitt - DLPFC - Sex"
author: "Kelsey Montgomery, Gabriel Hoffman, Thanneer Perumal"
date: "`r date()`"
output: html_document
---

@@

knit2synapse::createAndKnitToFolderEntity(file = "individual-cohorts/MSSM-Penn-Pitt_DLPFC_DE.Rmd",
                                 parentId ="syn22043922",
                                 folderName = "Differential Expression - Sex")
```

@@

```{r synapse.parameters, include=FALSE, cache=FALSE}
parentId = 'syn22043922';
activityName = 'Differential Expression';
activityDescription = 'Differential Expression - Sex';

@@

COVARIATES <- dplyr::rename(COVARIATES, Sex = Reported_Gender)

# Set covariates to adjust for from model selection
adjust.covars <- c("Sex", "ageOfDeath", "RIN2", "RIN", "IntronicRate", "IntragenicRate")
random_effect <- c("Institution")
```

### Normalisation - Sex
1. Sex is chosen as the primary variable of interest (i.e., covariate adjustments is conditioned on diagnosis)

@@

### Residual calculation
Calculate weighted residuals and add back "Sex" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Sex", colnames(ADJUSTED.FIT$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(ADJUSTED.FIT$design[,varsToAddIn])
```

@@

```{r diffExp, fig.height=10, fig.width=15}
L = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("SexFemale", "SexMale"))
contrast <- c(glue::glue_collapse(names(L[L != 0]), "-"))

@@

# Get differential expression
DE = lapply(glue::glue("L1"), function(i, FIT){
  topTable(FIT, coef=i, number = Inf) %>%
    rownameToFirstColumn('gene_id') %>% 
    left_join(backgroundGenes)
}, FIT.CONTR) 
names(DE) <- contrast

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('Sex','',Comparison),
                Direction = logFC/abs(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, into = c('ref.state','to.state'), sep = '-') %>%
  tidyr::unite(Comparison, `ref.state`, `to.state`, sep = '_vs_') %>%
  left_join(Ensemble2HGNC) %>%
  left_join(gene_lengths)
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

@@

# Code
CODE <- Folder(name = "Differential Expression - Sex", parentId = parentId)
CODE <- synStore(CODE)

# Store residual gene expression for network analysis
RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_DLPFC_Sex_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
nEXP_OBJ = File('CMC_DLPFC_Sex_netResidualExpression.tsv', 
                name = 'Sex Normalised, covariates removed residual expression (for network analysis)', 
                parentId = CODE$properties$id)
nEXP_OBJ = synStore(nEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(nEXP_OBJ, annotations = all.annotations)

# Store differential expression results
write.table(DE, file = 'CMC_DLPFC_Sex_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('CMC_DLPFC_Sex_DiffExpression.tsv', 
                name = 'Differential Expression Results (Sex)', 
                parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(DEXP_OBJ, annotations = all.annotations)

# Store adjusted fit linear mixed model object as RDS - Sex
saveRDS(ADJUSTED.FIT, file = 'CMC_DLPFC_Sex_dream-output.rds')
ADJ_OBJ = File('CMC_DLPFC_Sex_dream-output.rds',
               name = 'Dream Output object - Sex', 
               parentId = CODE$properties$id)
ADJ_OBJ = synStore(ADJ_OBJ, used = ALL_USED_IDs, activityName = activityName,
                   executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(ADJ_OBJ, annotations = all.annotations)
```
